---
title: "Somascan_Analysis"
format: html
editor: visual
---

# Read in the adat file and manipulate the dataframe

```{r}
library("SomaDataIO")

master <- read_adat("~/Documents/R scripts/Soma_project/L0125004423_v5.0_EDTAPlasma.hybNorm.medNormInt.plateScale.calibrate.anmlQC.qcCheck.anmlSMP.20251008.adat")

head(master)

```

## Manipulate the columns

```{r}
library(dplyr)

#remove calibrators and flagged samples
df <- master %>%
  filter(
    !SampleType %in% c("Calibrator", "Buffer", "QC"),
    RowCheck != "FLAG"
  ) %>% as_tibble()

#remove unwanted columns

df <- df %>% dplyr::select(-PlateId,-PlateRunDate, -ScannerID, -PlatePosition, -SlideId, -Subarray,-PercentDilution,-SampleMatrix,-Barcode,-Barcode2d,-SampleNotes,-AliquotingNotes, -SampleDescription, -AssayNotes, -ExtIdentifier, -SsfExtId,-SampleGroup,-SubjectID, -CLI, -RMA, -HybControlNormScale,-NormScale_20, -NormScale_0_005, -NormScale_0_5, -ANMLFractionUsed_20, -ANMLFractionUsed_0_005, -ANMLFractionUsed_0_5,-SiteId,-TimePoint, -SampleName, -SampleType, -RowCheck)

rownames(df) = NULL

```

## convert seq IDs to Uniprot

```{r}
library(stringr)
library(readr)
library(dplyr)

#load in the conversion key csv
map <- read.csv("seqid_to_uniprot.csv")

#reformatting (eg. 10023-32 to seq.10023.32)
map <- map %>% mutate(
  seqid_formatted = paste0("seq.", gsub("-",".",SeqId))
)

# make the vector from the map df
id_map <- setNames(map$UniProtID, map$seqid_formatted)

# don't rename the seqID column
old_names <- colnames(df)
new_names <- old_names
new_names[-1] <- ifelse(
  old_names[-1] %in% names(id_map),
  id_map[old_names[-1]],
  old_names[-1]
)

colnames(df) <- new_names

#remove any columns that still start with "seq" because these are controls

names(df) <- make.unique(names(df)) # some somamer sequences map to the same uniprotID but different regimes of the protein(cytoplasmic vs extracellular, etc. this makes the column names unique to avoid errors on dplyr.)

df <- df %>% dplyr::select(-starts_with("seq"))

```

## Add a Timepoint and Subject ID column

```{r}
df <- df %>%
  mutate(
    Timepoint = str_extract(SampleId, "D[1-5]$"),       # extract D1–D5 at the end
    SubjectID = str_remove(SampleId, "D[1-5]$")         # remove D1–D5 to get the subject ID
  ) %>%
  relocate(SubjectID, .before = SampleId) %>%            # move SubjectID before SampleId
  relocate(Timepoint, .after = SampleId)

df$Timepoint <- factor(df$Timepoint)

df <- as_tibble(df)
```

## log2 transformation of the proteomics data

```{r}

df_log2 <- df %>% mutate(across(where(is.numeric), ~ log2(.x)))

```

# Plot analytes of interest

```{r}
library(tidyverse)

# Vector of target proteins

# Inflammatory phase
# 
# target_proteins <- c("P01584",
# "P05231",
# "P01375",
# "P01579")


# Repair proteins
# 
#  target_proteins <- c("P01137",
#  "P15692",
#  "Q15063",
#  "P18075",
#  "P12643")

# Remodeling phase
# target_proteins <- c("P35222",
#                      "Q15063",
#                      "P04085",
#                      "P10451",
#                      "P01270")

# collagen peptides
#target_proteins <- c("Q03692",
                     # "P02452",
                     # "P02461",
                     # "Q15113",
                     # "Q9UKZ9")

 target_proteins <- c("P81172",
"Q92859",
"Q6ZVN8",
"Q15796",
"P36896",
"Q13253")

# Convert to long format for ggplot
df_long <- df %>%
  pivot_longer(cols = all_of(target_proteins),
               names_to = "Protein",
               values_to = "Expression")

# Summary stats: mean and SE per group
summary_df <- df_long %>%
  group_by(Protein, Timepoint) %>%
  summarise(
    mean_expr = mean(Expression, na.rm = TRUE),
    sd_expr = sd(Expression, na.rm = TRUE),
    .groups = "drop"
  )

# Plot
plot1 <- ggplot() + 
  geom_jitter(data = df_long, aes(x = Timepoint, y = Expression, color = Timepoint), width = 0.15, alpha = 0.6, size = 2) +
  geom_point(data = summary_df, aes(x = Timepoint, y = mean_expr), size = 3) + 
  facet_wrap(~ Protein, scales = "free_y", ncol = 4) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 13, margin = margin(b=20)),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  labs(
    title = "",
    x = "Draw",
    y = "Mean Expression"
  )
```

# tune sPLSDA keepX

```{r}
library(mixOmics)
library(dplyr)

set.seed(123)

X  <- df_log2 %>% dplyr::select(where(is.numeric)) %>% as.matrix()
Y  <- factor(df_log2$Timepoint)
ID <- df_log2$SubjectID

test_keepX <- c(5, 10, 25, 50, 75, 100, 150)
ncomp <- 3

tuned <- tune.splsda(
  X, Y,
  ncomp = ncomp,
  test.keepX = test_keepX,
  validation = "Mfold",
  folds = 5,          # <-- let mixOmics handle folds
  nrepeat = 20,       # repeated 5-fold CV
  measure = "BER",
  dist = "max.dist",
  progressBar = TRUE
)

tuned$choice.keepX

```


# Plot the sPLSDA with tuned keepX

```{r}
library(mixOmics)
library(tidyverse)

set.seed(123)
plsda_matrix <- df_log2 %>% dplyr::select((where(is.numeric)))
plsda_response <- df_log2$Timepoint
# run the PLSDA

plsda_result <- splsda(plsda_matrix, plsda_response, ncomp = 3, keepX = c(50,75,10), scale = T) #z scores the data within the plsda
PLS_scores <- as.data.frame(plsda_result$variates$X)
PLS_scores$Response <- df_log2$Timepoint

# plot the result

explained_var <- plsda_result$prop_expl_var$X
x_lab <- paste0("Component 1 (", round(explained_var[1] * 100, 1), "%)")
y_lab <- paste0("Component 2 (", round(explained_var[2] * 100, 1), "%)")


plsda_plot <- ggplot(PLS_scores, aes(x = comp1, y = comp2, color = Response, fill = Response)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(type = "norm", level = 0.85, geom = "polygon", alpha = 0.2, color = NA)+
  scale_fill_discrete(guide = "none") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Proteins over Time",
    x = x_lab,
    y = y_lab,
    color = "Blood Draw Timepoint",
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "right",
    axis.text = element_text(size = 14)
  )

loadings <- as.data.frame(plsda_result$loadings$X) %>%
  rownames_to_column(var="Protein")

top20_loadings <- loadings %>%
  mutate(abs_loading = abs(comp1)) %>%
  arrange(desc(abs_loading)) %>%
  dplyr::slice(1:20)

loadings_plot <- ggplot(top20_loadings, aes(x = reorder(Protein, comp1), y = comp1, fill = comp1 > 0)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("indianred3", "steelblue")) +
  coord_flip() +
  labs(
    title = "Top 20 PLS Loadings (C2)",
    x = "Protein",
    y = "Loading Value"
  ) +
  theme_minimal()

```


## Stratified Bootstrapping for feature stability analysis

```{r}
library(mixOmics)
library(tidyverse)

set.seed(123)

# this performs stratified resampling (subject-level bootstrapping), stores the variables from each component in each iteration, and then reports the frequencies of the variables in each model. Finally it selects variables per component whose frequency is > 0.4.


X <- df_log2 %>% dplyr::select(where(is.numeric)) %>% as.matrix() # define the matrix
Y <- df_log2$Timepoint
SubjectID <- df_log2$SubjectID

nboot <- 500 # number of iterations
ncomp <- 3
keepX <- c(50, 75, 10)


# cluster bootstrap: sample SUBJECTS with replacement; keep all rows per subject
cluster_boot_idx <- function(subject_id) {
  subj <- unique(subject_id)
  boot_subj <- sample(subj, length(subj), replace = TRUE)
  unlist(lapply(boot_subj, function(s) which(subject_id == s)), use.names = FALSE)
}

# store selected variables per component across bootstraps
selected_vars <- vector("list", ncomp)
for (k in seq_len(ncomp)) selected_vars[[k]] <- character(0)

# ---- bootstrap loop ----
for (b in seq_len(nboot)) {

  idx <- cluster_boot_idx(SubjectID)

  fit <- splsda(
    X[idx, ],
    Y[idx],
    ncomp = ncomp,
    keepX = keepX,
    scale = TRUE
  )

  for (k in seq_len(ncomp)) {
    selected_vars[[k]] <- c(selected_vars[[k]], selectVar(fit, comp = k)$name)
  }
}

# ---- variables per component with selection frequency ----
freq_tables <- lapply(seq_len(ncomp), function(k) {
  tibble(Protein = selected_vars[[k]]) %>%
    count(Protein) %>%
    mutate(freq = n / nboot) %>%
    arrange(desc(freq))
})
names(freq_tables) <- paste0("Comp", seq_len(ncomp))

stable_vars_04 <- lapply(freq_tables, function(df) {
  df %>%
    filter(freq > 0.4) %>%
    pull(Protein)
})

```


### pull stable features from components 1 and 2 and convert to entrezID

```{r}
library(AnnotationDbi)
library(org.Hs.eg.db)

# Pull the STABLE features from each component for ORA.

# Step 1. stable features for component 1 (freq > 0.4)
stable_c1 <- stable_vars_04$Comp1
stable_c2 <- stable_vars_04$Comp2

# Step 2. pull the loadings and variable names for all 75 variables in component 1.

c1_loadings <- plsda_result$loadings$X[,1]

# Step 4. Split them into the positive and negative components for interpretability down the line (to know which pathways are up and which are down)

c1_pos_uniprot_ids <- names(c1_loadings)[c1_loadings > 0]
c1_neg_uniprot_ids <- names(c1_loadings)[c1_loadings < 0]


# Step 5. compare the uniprot IDs to the list of stable features, then remove the suffixes.
c1_pos_uniprot_ids_stable <- intersect(c1_pos_uniprot_ids, stable_c1)
c1_neg_uniprot_ids_stable <- intersect(c1_neg_uniprot_ids, stable_c1)

c1_pos_uniprot_ids_stable <- c1_pos_uniprot_ids_stable %>% str_replace("\\.\\d+$","")
c1_neg_uniprot_ids <- c1_neg_uniprot_ids_stable %>% str_replace("\\.\\d+$","")

# Map UniProt -> Entrez
pls_pos_map <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = unique(c1_pos_uniprot_ids_stable),
  keytype = "UNIPROT",
  columns = c("ENTREZID")
) %>%
  filter(!is.na(ENTREZID)) %>%
  distinct(UNIPROT, ENTREZID)

c1_pos_gene_ids <- unique(pls_pos_map$ENTREZID)

pls_neg_map <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = unique(c1_neg_uniprot_ids_stable),
  keytype = "UNIPROT",
  columns = c("ENTREZID")
) %>%
  filter(!is.na(ENTREZID)) %>%
  distinct(UNIPROT, ENTREZID)

c1_neg_gene_ids <- unique(pls_neg_map$ENTREZID)

# Repeat for component 2

c2_loadings <- plsda_result$loadings$X[,2]
c2_pos_uniprot_ids <- names(c2_loadings)[c2_loadings > 0]
c2_neg_uniprot_ids <- names(c2_loadings)[c2_loadings < 0]

c2_pos_uniprot_ids_stable <- intersect(c2_pos_uniprot_ids, stable_c2)
c2_neg_uniprot_ids_stable <- intersect(c2_neg_uniprot_ids, stable_c2)

c2_pos_uniprot_ids_stable <- c2_pos_uniprot_ids_stable %>% str_replace("\\.\\d+$","")
c2_neg_uniprot_ids_stable <- c2_neg_uniprot_ids_stable %>% str_replace("\\.\\d+$","")

pls_pos_map2 <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = unique(c2_pos_uniprot_ids_stable),
  keytype = "UNIPROT",
  columns = c("ENTREZID")
) %>%
  filter(!is.na(ENTREZID)) %>%
  distinct(UNIPROT, ENTREZID)

c2_pos_gene_ids <- unique(pls_pos_map2$ENTREZID)

pls_neg_map2 <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = unique(c2_neg_uniprot_ids_stable),
  keytype = "UNIPROT",
  columns = c("ENTREZID")
) %>%
  filter(!is.na(ENTREZID)) %>%
  distinct(UNIPROT, ENTREZID)

c2_neg_gene_ids <- unique(pls_neg_map2$ENTREZID)

```

# GSEA on all timepoints compared to Draw 1

## LIMMA for DEGs and rank vectors

```{r}
library(limma)
library(dplyr)

# Create design matrix for timepoint comparison

design <- model.matrix(~ 0 + Timepoint, data = df_log2)
colnames(design) <- levels(df_log2$Timepoint)
expr_matrix <- df_log2 %>% dplyr::select((where(is.numeric))) %>% as.matrix()
expr_matrix_transpose <- expr_matrix %>% t() # transpose the matrix. I found I had to do this.

contrast_matrix <- makeContrasts(
  D1_v_D2 = D1 - D2,
  D2_v_D3 = D2 - D3,
  D3_v_D4 = D3 - D4,
  D4_v_D5 = D4 - D5,
  D3_v_D5 = D5 - D3,
  D1_v_D3 = D1 - D3,
  D1_v_D4 = D1 - D4,
  D1_v_D5 = D1 - D5,
  custom = (D1+D2)/2 - (D3 + D4 + D5)/3,
  levels = design
)

# Estimate Sample Correlation
SubjectID <- df_log2$SubjectID
corfit <- duplicateCorrelation(expr_matrix_transpose, design, block = SubjectID)

# Fit the model
fit <- lmFit(expr_matrix_transpose, design, block = SubjectID, correlation=corfit$consensus)
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2) #Bayesian noise correction

#Find the top proteins for each contrast. By default, this topTable function uses BH FDR correction. I kept changing number until the significant features were all captured.
D1_v_D2 <- topTable(fit2, coef = "D1_v_D2", number = 1500)
D1_v_D3 <- topTable(fit2, coef = "D1_v_D3", number = 1500)
D1_v_D4 <- topTable(fit2, coef = "D1_v_D4", number = 1500)
D1_v_D5 <- topTable(fit2, coef = "D1_v_D5", number = 1500)
custom <- topTable(fit2, coef = "custom", number = 1500)

# extras
D2_v_D3 <- topTable(fit2, coef = "D2_v_D3", number = 1500)
D3_v_D4 <- topTable(fit2, coef = "D3_v_D4", number = 1500)
D4_v_D5 <- topTable(fit2, coef = "D4_v_D5", number = 1500)
D3_v_D5 <- topTable(fit2, coef = "D3_v_D5", number = 1500)

# Find the significant features
D1_D2_sig <- D1_v_D2 %>% filter(adj.P.Val < 0.05)
D1_D3_sig <- D1_v_D3 %>% filter(adj.P.Val < 0.05)
D1_D4_sig <- D1_v_D4 %>% filter(adj.P.Val < 0.05)
D1_D5_sig <- D1_v_D5 %>% filter(adj.P.Val < 0.05)
custom_sig <- custom %>% filter(adj.P.Val < 0.05)
D2_D3_sig <- D2_v_D3 %>% filter(adj.P.Val < 0.05)
D3_D4_sig <- D3_v_D4 %>% filter(adj.P.Val < 0.05)
D4_D5_sig <- D4_v_D5 %>% filter(adj.P.Val < 0.05)
D3_D5_sig <- D3_v_D5 %>% filter(adj.P.Val < 0.05)

```

## Some formatting and conversions

```{r}
library(dplyr)
library(AnnotationDbi)
library(org.Hs.eg.db)

# strip probe suffix like ".1" or ".2" from UniProt IDs, choose the uniprot with higher signal, then convert to ensemble. do this for all 4 contrasts.

# D1 vs D2

D1_D2 <- D1_v_D2 %>%
  tibble::rownames_to_column(var = "Uniprot_ID") %>%
  mutate(Uniprot_ID = sub("\\.\\d+$", "", Uniprot_ID)) %>%
  group_by(Uniprot_ID) %>%
  summarise(stat = t[which.max(abs(t))], .groups = "drop") %>%  # collapse multiple probes per UniProt (keep strongest |t|)
  mutate(
    ENTREZID = AnnotationDbi::mapIds(
      org.Hs.eg.db,
      keys      = Uniprot_ID,
      keytype   = "UNIPROT",
      column    = "ENTREZID",
      multiVals = "first"   # or "filter" to drop ambiguous UniProts
    )
  ) %>%
  filter(!is.na(ENTREZID)) %>%
  group_by(ENTREZID) %>%
  summarise(stat = stat[which.max(abs(stat))], .groups = "drop") %>%   # collapse to one value per Entrez gene
  arrange(desc(stat))

# D1 vs D3

D1_D3 <- D1_v_D3 %>%
  tibble::rownames_to_column(var = "Uniprot_ID") %>%
  mutate(Uniprot_ID = sub("\\.\\d+$", "", Uniprot_ID)) %>%
  group_by(Uniprot_ID) %>%
  summarise(stat = t[which.max(abs(t))], .groups = "drop") %>%  # collapse multiple probes per UniProt (keep strongest |t|)
  mutate(
    ENTREZID = AnnotationDbi::mapIds(
      org.Hs.eg.db,
      keys      = Uniprot_ID,
      keytype   = "UNIPROT",
      column    = "ENTREZID",
      multiVals = "first"   # or "filter" to drop ambiguous UniProts
    )
  ) %>%
  filter(!is.na(ENTREZID)) %>%
  group_by(ENTREZID) %>%
  summarise(stat = stat[which.max(abs(stat))], .groups = "drop") %>%   # collapse to one value per Entrez gene
  arrange(desc(stat))

# D1 vs D4

D1_D4 <- D1_v_D4 %>%
 tibble::rownames_to_column(var = "Uniprot_ID") %>%
  mutate(Uniprot_ID = sub("\\.\\d+$", "", Uniprot_ID)) %>%
  group_by(Uniprot_ID) %>%
  summarise(stat = t[which.max(abs(t))], .groups = "drop") %>%  # collapse multiple probes per UniProt (keep strongest |t|)
  mutate(
    ENTREZID = AnnotationDbi::mapIds(
      org.Hs.eg.db,
      keys      = Uniprot_ID,
      keytype   = "UNIPROT",
      column    = "ENTREZID",
      multiVals = "first"   # or "filter" to drop ambiguous UniProts
    )
  ) %>%
  filter(!is.na(ENTREZID)) %>%
  group_by(ENTREZID) %>%
  summarise(stat = stat[which.max(abs(stat))], .groups = "drop") %>%   # collapse to one value per Entrez gene
  arrange(desc(stat))

# D1 vs D5

D1_D5 <- D1_v_D5 %>%
  tibble::rownames_to_column(var = "Uniprot_ID") %>%
  mutate(Uniprot_ID = sub("\\.\\d+$", "", Uniprot_ID)) %>%
  group_by(Uniprot_ID) %>%
  summarise(stat = t[which.max(abs(t))], .groups = "drop") %>%  # collapse multiple probes per UniProt (keep strongest |t|)
  mutate(
    ENTREZID = AnnotationDbi::mapIds(
      org.Hs.eg.db,
      keys      = Uniprot_ID,
      keytype   = "UNIPROT",
      column    = "ENTREZID",
      multiVals = "first"   # or "filter" to drop ambiguous UniProts
    )
  ) %>%
  filter(!is.na(ENTREZID)) %>%
  group_by(ENTREZID) %>%
  summarise(stat = stat[which.max(abs(stat))], .groups = "drop") %>%   # collapse to one value per Entrez gene
  arrange(desc(stat))

# D2 vs D3

custom_tib <- custom %>%
  tibble::rownames_to_column(var = "Uniprot_ID") %>%
  mutate(Uniprot_ID = sub("\\.\\d+$", "", Uniprot_ID)) %>%
  group_by(Uniprot_ID) %>%
  summarise(stat = t[which.max(abs(t))], .groups = "drop") %>%  # collapse multiple probes per UniProt (keep strongest |t|)
  mutate(
    ENTREZID = AnnotationDbi::mapIds(
      org.Hs.eg.db,
      keys      = Uniprot_ID,
      keytype   = "UNIPROT",
      column    = "ENTREZID",
      multiVals = "first"   # or "filter" to drop ambiguous UniProts
    )
  ) %>%
  filter(!is.na(ENTREZID)) %>%
  group_by(ENTREZID) %>%
  summarise(stat = stat[which.max(abs(stat))], .groups = "drop") %>%   # collapse to one value per Entrez gene
  arrange(desc(stat))
```


### Using a priori chosen pathways

Import the pathways of interest from Reactome (these were downloaded directly from reactome website)

```{r}
library(tidyverse)
library(stringr)


pathway_dir <- "reactome_pathways"
files <- list.files(pathway_dir, pattern = "\\.tsv$", full.names = TRUE)

# I manually renamed these to include the pathway name. it takes a long time but i could not think of a better way to do it. Much of the below is chatgpt script that I need to spend time understanding.

term2gene <- purrr::map_dfr(files, function(f) {
  term <- str_extract(basename(f), "R-HSA-\\d+")
  
  df <- readr::read_tsv(f, show_col_types = FALSE) %>%
    # standardize header names
    rename_with(tolower) 
  
  df2 <- df %>%
    transmute(
      term = term,
      uniprot = .data[["identifier"]]   # UniProt column
    )
  
  df2
}) %>%
  filter(!is.na(uniprot), uniprot != "", uniprot != "identifier") %>%
  distinct(term, uniprot)

head(term2gene)

term2name <- map_dfr(files, parse_filename) %>%
  distinct(term, name)
```

Format as Entrez ID (since other databases expect entrez and my ranks are already in entrez from previous iterations)

```{r}

library(dplyr)
library(clusterProfiler)
library(org.Hs.eg.db)

t2g_entrez <- bitr(
  term2gene$uniprot,
  fromType = "UNIPROT",
  toType   = "ENTREZID",
  OrgDb    = org.Hs.eg.db
) %>%
  distinct(UNIPROT, ENTREZID)

term2gene_entrez <- term2gene %>%
  inner_join(t2g_entrez, by = c("uniprot" = "UNIPROT")) %>%
  transmute(term = term, gene = ENTREZID) %>%
  distinct(term, gene)

head(term2gene_entrez)

```
Run GSEA using clusterprofiler

```{r}

library(clusterProfiler)
library(tidyverse)
library(enrichplot)

ranks <- D1_D2$stat
names(ranks) <- D1_D2$ENTREZID
ranks <- sort(ranks, decreasing = TRUE)

gsea_custom <- GSEA(
  geneList      = ranks,
  TERM2GENE     = term2gene_entrez,
  TERM2NAME     = term2name,
  minGSSize     = 10,
  maxGSSize     = 300,
  pAdjustMethod = "none",
  verbose       = FALSE
)

gsea_custom@result <- gsea_custom@result %>%
  mutate(.sign = ifelse(NES >= 0, "Positive", "Negative"))

dotplot <- dotplot(gsea_custom, showCategory = 10, split = ".sign") +
  facet_grid(. ~ .sign)

```


